Equivalence Testing and the Second Generation P-Value

### 2018-07-28 sgpv_vs_TOST.Rmd
fix typo

#The relationship between *p*values from TOST and SGPV

@@

  invisible(capture.output(res <- TOSTone.raw(m = m, 
                                              mu = mu,
                                              sd = sd, 
                                              n = n, 
                                              low_eqbound = low_eqbound, 
                                              high_eqbound = high_eqbound, 
                                              alpha = alpha,
                                              plot = FALSE
  )))
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 5*: The relationship between *p*values from the TOST procedure and the SGPV for the same scenario as in Figure 1.

@@

We can again plot the two statistics against each other so see where they are unrelated (indicated by staight lines in the curve). We see the SGPV is 0.5 for a range of observed means where the p-value from the equivalence test still varies. It should be noted that in these calculations the *p*-values for the TOST procedure are *never* smaller than 0.05 (i.e., they do not get below 0.05 on the y-axis). In other words, we can't conclude equivalence based on any of the observed means. How is this possible? Remember that the TOST procedure consists of two one-sided tests against the upper and lower equivalence bound. The TOST *p*-value is smaller than 0.05 if the 90% CI falls completely between the upper and lower equivalence bounds. However, we are examining a scenario where the 90% CI is so wide that it never falls completely within the two equivalence bounds. As Lakens (2017) notes: “in small samples (where CIs are wide), a study might have no statistical power (i.e., the CI will always be so wide that it is necessarily wider than the equivalence bounds).” None of the *p*-values based on the TOST procedure are below 0.05, and thus, in the long run we have 0% power.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 7*: The relationship between *p*values from the TOST procedure and the SGPV for the same scenario as in Figure 6.

@@

If the observed mean would be somewhat closer to 0, or further away from 0, the SGPV would remain constant (the CI width does not change, it ccompletely overlaps with the equivalence range) while the *p*-value for the TOST procedure varies. We can see this in Figure 9 below. The SGPV is not set to 0.5, but is slightly higher than 0.5 across a range of means. How high the SGPV will be when the CI overlaps with the lower and upper equivalence bounds, but the CI is not twice as large as the equivalence range, depends on the width of the CI and the equivalence range. 
```{r, include=FALSE}
step = 0.01

@@

If we once more plot the two statistics against each other so see where they are unrelated (indicated by staight lines in the curve). We see the SGPV is 0.56 for a range of observed means where the *p*-value from the equivalence test still varies.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 10*: The relationship between *p*values from the TOST procedure and the SGPV for the same scenario as in Figure 6.

To conclude this section, there are four situations where the *p*-value from the TOST procedure is unrelated to the SGPV, In all these situation the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely descriptive statistic, the SGPV is more limited than the *p*value from the TOST procedure. The proportion of overlap can be the same value when the observed mean is 0 or when the observed mean falls just inside the equivalence bound, and additional information (e.g., the 95% CI) is required to differentiate these situations. One way to mitigate this limitation of the SGPV would be to set the SGPV to 0.5 whenever the CI overlaps with both the upper and lower equivalence bound (irrespective of the width of the CI).

@@

It seems Blume et al (2018) where not aware of the existence of equivalence tests, and we believe that our explanation of the similarities between the TOST procedure and the SGPV provides some useful context to interpret the contribution of second generation *p*values to the statistical toolbox. The novelty lies in its use as a descriptive statistic. The added benefit of calculating the proportion of overlap of a 95% CI with the equivalence range, and using this percentage to describe the data, remains somewhat unclear for practical puroposes. Nevertheless, our only goal is to clarify the relationship between a newly proposed statistic and the already existing TOST approach used to test for equivalence, and let researchers make an informed decision about which statistical approach provides the best answer to their question.

>>>>

#The relationship between *p*-values from TOST and SGPV

@@

  invisible(capture.output(res <- TOSTone.raw(m = m, 
                                              mu = mu,
                                              sd = sd, 
                                              n = n, 
                                              low_eqbound = low_eqbound, 
                                              high_eqbound = high_eqbound, 
                                              alpha = alpha,
                                              plot = FALSE
  )))
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 5*: The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure 1.

@@

We can again plot the two statistics against each other so see where they are unrelated (indicated by staight lines in the curve). We see the SGPV is 0.5 for a range of observed means where the p-value from the equivalence test still varies. It should be noted that in these calculations the *p*-values for the TOST procedure are *never* smaller than 0.05 (i.e., they do not get below 0.05 on the y-axis). In other words, we can't conclude equivalence based on any of the observed means. How is this possible? Remember that the TOST procedure consists of two one-sided tests against the upper and lower equivalence bound. The TOST *p*-value is smaller than 0.05 if the 90% CI falls completely between the upper and lower equivalence bounds. However, we are examining a scenario where the 90% CI is so wide that it never falls completely within the two equivalence bounds. As Lakens (2017) notes: “in small samples (where CIs are wide), a study might have no statistical power (i.e., the CI will always be so wide that it is necessarily wider than the equivalence bounds).” None of the *p*-values based on the TOST procedure are below 0.05, and thus, in the long run we have 0% power.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 7*: The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure 6.

@@

If the observed mean would be somewhat closer to 0, or further away from 0, the SGPV would remain constant (the CI width does not change, it completely overlaps with the equivalence range) while the *p*-value for the TOST procedure varies. We can see this in Figure 9 below. The SGPV is not set to 0.5, but is slightly higher than 0.5 across a range of means. How high the SGPV will be when the CI overlaps with the lower and upper equivalence bounds, but the CI is not twice as large as the equivalence range, depends on the width of the CI and the equivalence range. 
```{r, include=FALSE}
step = 0.01

@@

If we once more plot the two statistics against each other so see where they are unrelated (indicated by staight lines in the curve). We see the SGPV is 0.56 for a range of observed means where the *p*-value from the equivalence test still varies.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 10*: The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure 6.

To conclude this section, there are four situations where the *p*-value from the TOST procedure is unrelated to the SGPV, In all these situation the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely descriptive statistic, the SGPV is more limited than the *p*-value from the TOST procedure. The proportion of overlap can be the same value when the observed mean is 0 or when the observed mean falls just inside the equivalence bound, and additional information (e.g., the 95% CI) is required to differentiate these situations. One way to mitigate this limitation of the SGPV would be to set the SGPV to 0.5 whenever the CI overlaps with both the upper and lower equivalence bound (irrespective of the width of the CI).

@@

It seems Blume et al (2018) where not aware of the existence of equivalence tests, and we believe that our explanation of the similarities between the TOST procedure and the SGPV provides some useful context to interpret the contribution of second generation *p*-values to the statistical toolbox. The novelty lies in its use as a descriptive statistic. The added benefit of calculating the proportion of overlap of a 95% CI with the equivalence range, and using this percentage to describe the data, remains somewhat unclear for practical puroposes. Nevertheless, our only goal is to clarify the relationship between a newly proposed statistic and the already existing TOST approach used to test for equivalence, and let researchers make an informed decision about which statistical approach provides the best answer to their question.
